---
title: "Module_02"
author: "Justin Proctor"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# MB5370 Module 2: R for Marine Science

# Workshop 1: Foundations of Data Science — Wrangling and Plotting

## Introduction
This rmd file covers all the material for the first R for Marine Science workshop, which focuses on foundational data science skills for marine scientists. We will cover the following topics:
- Importing data from various file formats
- Exploring and summarizing data structures
- Wrangling data with the tidyverse framework

### Clear out environment
```{r check objects in environment}
## Inventory every active object currently residing in session RAM
objects()
```

```{r housekeeping, message=FALSE, warning=FALSE, include=FALSE}
## Purge global environment
rm(list=ls())
# Confirm that global session memory now completely vacant
objects()
```

```{r}
#install.packages("here")
```


### Load packages
```{r Load the primary data science framework and Excel import library}
library(tidyverse)
library(readxl)
library(here)
```

```{r practice imports}
# Practice Import A: Loading a standard comma-separated plain text file
benthic_cover <- read_csv(here::here("data/reef_cover_log.csv"))

# Practice Import B: Parsing a tab-separated telemetry instrument array string
acoustic_stream <- read_tsv(here::here("data/acoustic_telemetry_stream.txt"))

# Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet
fisheries_annual <- read_excel(here::here("data/fish_catch_data.xlsx"), sheet = "Commercial_2026")

```
### Workflow
```{r loading in dataset}
# Read in mangrove_data
mangrove_data <- read_csv(file = here::here("data/mangrove_survey_raw.csv"))
```

```{r troublshooting for mangrove data}
# Use args within read_csv to skip headers and declare missing flags
mangrove_data <- read_csv(
  here::here("data/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA
```

```{r visualizing how two data structures behave differently under pressure}
# Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)
```

```{r visualizing how two data structures behave differently under pressure}
# Print the old-style dataframe structure to view
print(benthic_cover_df)
# And compare with tibble alternative
print(benthic_cover)
```
```{r load and assign data from palmerpenguins package}
# Load the package data into active memory
library(palmerpenguins)
data("penguins")
```

```{r Examine the structure of the dataset}
#always do this when loading a new dataset!
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version
```

```{r Generate an exploratory summary matrix}
summary(penguins)
```

```{r isolating or dropping columns based on variable names}
# Vertically slice specific morphometric variables by explicit name
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)

# Discard logistics tracking attributes while preserving everything else using the minus sign
clean_scientific_fields <- select(penguins, -year)
```

```{r Sifting rows with filter()}
# Isolate observations belonging to a single categorical target group
adelie_cohort <- filter(penguins, species == "Adelie")

# Sift out individuals using continuous numerical boundary thresholds
# Preserves only large penguins whose mass exceeds 4500 grams
heavy_penguins <- filter(penguins, body_mass_g > 4500)

# Combine multiple conditional parameters across separate attributes
# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

# Sift records matching multiple targeting flags within an explicit set
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
```

```{r Ordering sequences with arrange}
# essential for inspecting size hierarchies or chronologically ordering long-term environmental observation linesessential for inspecting size hierarchies or chronologically ordering long-term environmental observation linesessential for inspecting size hierarchies or chronologically ordering long-term environmental observation lines

# Sort penguins by ascending body mass (Default setting: Smallest mass first)
lightest_first <- arrange(penguins, body_mass_g)

# Sort penguins in descending sequence using the desc() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))

# Execute nested sorting criteria: Group by species, then sort by descending bill length
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
```

```{r using the pipe}
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
```

```{r Computing new attributes with mutate}
# we use the mutate() verb to modify existing attributes or append entirely new vectors to the data frame
# Calculate a new morphological ratio in our environment
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )

# View your newly engineered variables appended to the far-right columns
glimpse(penguin_ratios)
```

```{r Data aggregation and ecological summarisation}
# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)
```

```{r Overcoming the missing value trap}
# If a column contains even a single missing observation flag (NA), any mathematical aggregation function (such as mean(), sd(), or sum()) will automatically return NA to protect you from miscalculating metrics on incomplete data. To override this safely, you must explicitly declare that missing cells should be dropped during calculation
# Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)

```

```{r Challenge: Visualising Body Size and Shape of penguins}
# Summary table
penguins %>%
drop_na(species, island, body_mass_g) %>%
group_by(species, island) %>%
summarise(
mean_body_mass = mean(body_mass_g),
.groups = "drop"
)

# Boxplot
penguins %>%
drop_na(species, island, body_mass_g) %>%
ggplot(aes(x = species, y = body_mass_g, fill = island)) +
geom_boxplot() +
labs(
title = "Body Mass by Species and Island",
x = "Species",
y = "Body Mass (g)"
)
```

```{r Example: Visualising Mean Trends with Uncertainty}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot
```

```{r challenge: visualizing mean trends using SE instead of SD}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
     se_mass = sd_mass / sqrt(n),  # Calculate standard error
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = mean_mass - se_mass, 
                    ymax = mean_mass + se_mass), 
                width = 0.2,  position = position_dodge(width = 0.4)) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard SE",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot
```

```{r create folders for saving and exporting}
# Create output directories if they do not exist:
if (!dir.exists("../output/figures")) dir.create("../output/figures") # folder for figs
if (!dir.exists("../output/tables")) dir.create("../output/tables") # folder for tables
if (!dir.exists("../Rdata")) dir.create("../Rdata") # folder for Rdata objects
```

```{r saving table}
# 1. Exporting our collapsed summary table as a universal flat text file
write_csv(biological_signal, "../output/tables/penguin_species_mass_summary.csv")

# 2. Saving our cleaned morphological cohort table as a native R binary file
saveRDS(adelie_cohort, "../output/tables/clean_penguin_morphology_cohort.rds")
```

```{r saving figure}
ggsave("../output/figures/mass_compare_plot.png", 
       plot = mass_compare_plot, 
       width = 120, height = 120, 
       units = "mm", dpi = 300)
```


# Workshop 2: Advanced data wrangling 
## Extracting ecological signals from noisy systems

## Introduction
This rmd file covers all the material for the first R for Marine Science workshop. 
Here we are going to cover:
- Tidying data
- Pivoting wider and longer
- Separating and uniting variables
- Dealing with stings and factors
- Handling missing data values

### Workflow
```{r}
library(tidyverse)
```

```{r load table1}
table1
```

```{r compute rate per 10k}
table1 %>% 
  mutate(rate = cases / population * 10000)
```

```{r compute cases per year}
table1 %>% 
  count(year, wt = cases)
```

```{r visualise case changes over time}
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

```{r installing billboard dataset}
#install.packages("billboard")
summary(billboard)
```

```{r adjusting format of billboard using pivot_longer}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )
```

```{r dropping NA values out of dataset when songs werent in the top 100}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE)
```

```{r creating df dataset to practice pivot longer}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
```

```{r pivoting longer on df dataset}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```

```{r installing package for pivot wider practice}
#install.packages("cms_patient_experience")
summary(cms_patient_experience)
```

```{r using distinct to see full set of measures in the dataset}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)
```

```{r applying pivot wider}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
```

```{r applying pivot wider with id_cols to preserve org identifiers}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

```{r}
df2 <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```


```{r further visualizing pivot wider}
df2 |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

```{r establishing row and column identifiers for pivot wider}
df2 |> 
  distinct(measurement) |> 
  pull()
```

```{r establishing row and column identifiers for pivot wider}
df2 |> 
  select(-measurement, -value) |> 
  distinct()
```

```{r combining results of previous two chunks to create a skeleton for pivot wider}
df2 |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
```

```{r using pivot longer on penguins data}
library(palmerpenguins)

# Create a long version of the penguins dataset

penguins_long <- penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# View the result
head(penguins_long)
```

```{r visualizing new lengthened format}
penguin_morphometric_plot <- penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )

ggsave("../output/penguin_morphometric_dist.png", 
       plot = penguin_morphometric_plot, 
       width = 120, height = 120, 
       units = "mm", dpi = 300)
```

```{r widening penguin data}
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary)
```

```{r creating clean matrix}
mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )

head(mass_matrix)
```

```{r load table}
table3
```

```{r separating rate into cases and population}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```

```{r rewriting the code above using a specific character to separate a column}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")
```

```{r convert to integer values}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```

```{r separating year into century and year - use this arrangement to separate the last two digits of each year.}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```

```{r uniting century and year back into a single column}
table5 %>% 
  unite(new, century, year, sep = "")
```

```{r Standardising text with stringr using a messy dataset}
# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Replace any spaces with underscores
    site_clean = str_trim(site_clean),
    # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_")
  )

print(clean_sites)
```

```{r using lubridate to align dates}
library(lubridate)

# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2
```

```{r parsing with exact timestamps}
# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)
```

```{r setting up two tables to demonstrate joining}
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)
```

```{r joining data}
# Joining metadata to our observations
joined_data <- observations |>
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)
```

```{r only keeping rows that have complete matches in both tables}
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
glimpse(matched_data)
```

```{r using an anti_join() to find out exactly which site codes are misspelled or missing from metadata table.}
# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code))
```

```{r using dplyr::na_if() to omit data-outputs where a concrete value actually represents a missing value}
# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data |>
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)
```

```{r Replacing missing values with coalesce()}
# Simulate some count data
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

# Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)
```

```{r identifying an NaN: value is mathematically impossible and indicates a division by zero or an undefined operation}
cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours)

print(cpue_calc)
```

```{r loading raw catch data to address NaN issue}
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))
```

```{r explicitly creating rows for the missing combinations, and allowing the resulting NAs to be filled with 0}
# Force inclusion of hidden zero-catch data
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)
```

```{r removing the entire row using drop_na() as a final/last resort option}
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

# Remove any row where salinity is missing
clean_log <- sensor_log |>
  drop_na(salinity)
```

### Exercise 1: Cleaning the messy metadata
```{r}
# Make sure packages are loaded
library(tidyverse)
library(lubridate)
library(palmerpenguins)

# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print(island_metadata)
```

```{r cleaned up dataset}
clean_metadata <- island_metadata |>
  mutate(
    # Clean island names: remove spaces + standardize case
    island_name = stringr::str_trim(island_name),
    island_name = stringr::str_to_title(island_name),

    # Convert to Date (handles mixed formats)
    station_install = lubridate::dmy(station_install)
  )

print(clean_metadata)
```

### Exercise 2: The Relational Join
```{r}
penguins_spatial <- penguins |>
  left_join(clean_metadata, by = join_by(island == island_name))

# Scroll to the right to see the new columns
head(penguins_spatial)
```

###Exercise 3: The Wide Summary Matrix
```{r}
max_mass_table <- penguins_spatial |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(
    max_body_mass_g = max(body_mass_g),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = island,
    values_from = max_body_mass_g
  )

max_mass_table
```
