---
title: "WKSP_02_KeystoneExercise"
author: "Justin Proctor"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: html
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

## Module 2: R for Marine Science

## Workshop 2: **Advanced Data Wrangling**

# **Keystone exercise: Estuary Fish Survey - Data Rescue Mission**

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

### Introduction

Here we are going to wrangle in messy data on the Ross River Estuary fish surveys and produce final clean tables and figure.

### Housekeeping
```{r housekeeping}
rm(list =ls())
```

### Packages
```{r packages}
library(tidyverse)
library(readxl)
```

### Load Files/Main Dataset
```{r load metadata/files}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
View(metadata)
```

```{r loading/creating main dataset}
catch1 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
```

```{r inspect individual tabs}
catch1
catch2
catch3
catch4
```

```{r bind rows together}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
rm(catch1, catch2, catch3, catch4) #clean up the environment
```

```{r alternative method for loading files}
file_path <- "../data/workshop2/estuary_catch_log.xlsx"
catch_data2 <- excel_sheets(file_path) |> 
  map_dfr(~read_excel(path = file_path, sheet = .x))
```

### Examine Data
```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

Can't tell from these outputs how many levels of site and species we have. Two options: let's check those explicitly by using distinct() or can convert to factors Let's look at the distinct values first, as we might need to clean before converting to factor vars.

```{r}
catch_data |>
  distinct(site)

catch_data |>
  distinct(species)
```

OK, so we have four species in there as well as four sites which look to be located at different points on the Ross River.
Need to:
- Fix site names and place in reasonable order. 
- Can decide this now or we can load the metadata and ensure consistency
- Fix species names - to do this we'll need to make consistent and then use the taxonomic dictionary to translate to scientific names.

### Load, Inspect, and Standardize Metadata
```{r inspect metadata}
head(metadata)
glimpse(metadata)
```

site names are all over the place in metadata - will have to standardize

```{r clean up site names}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

```{r recheck}
catch_data |> 
  str()
# catch_data |> 
#   distinct(site)
glimpse(metadata)
```

### Load Taxanomic Dictionary and Sensor Data
```{r load taxonimic dictionary}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
head(species_dict)
```

```{r load the sensor data}
sensor_row <- read_csv("../data/workshop2/estuary_sonde_data.csv")
```

```{r check sensor data}
head(sensor_row)
str(sensor_row)
glimpse(sensor_row)
summary(sensor_row)
```

### Wrangle Sonde Data
```{r celan up sensor data}
sensor_data <- sensor_row |> 
  mutate(
    # parse the messy character string (day/month/year hour:minute)
    datetime = dmy_hm(timestamp),
    # convert the -999.0 hardware error to a true NA
    turbidity = na_if (turbidity, -999.0),
    # convert negative salinity values to NA
    salinity = ifelse(salinity < 0, NA, salinity),
  )
```

```{r clean up daily sensor data}
sensor_daily <- 
  sensor_data |> 
  mutate(date = as_date(floor_date(datetime, "day"))) |>
  group_by(site, date) |> 
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    mean_turbidity = mean(turbidity, na.rm = TRUE),
    .groups = "drop"
  )
```

### Notes:
-   Environmental variables needed cleaning:
-   Turbidity data has a failed value (-999)
-   Salinity had some negative values which are not possible and should be converted to NA
-   Site ok

### Wrangle Catch Data
```{r standardize variable in the catch log to match dictionaries}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_"),
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```

```{r check common names in catch log match those in the taxonomic dictionary}
catch_data |> 
  distinct(species)
species_dict |>
  distinct(common_name)
```

```{r taxonomic translation (common to scientific)}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name)) |>
```

### Need to:

-   Join with metadata and sensor data
-   Implicit zero observations

4 x spp 4 x sites 30 surveys = 4*4*30

### Join Catch Data with Metadata and Sensor Data
```{r join catch data with metadata and sensor data}
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))
```

### Produce Final Table
```{r}
catch_data_final |>
  complete(
    nesting(site, date, lat, lon, zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    #convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )
```

### Produce summary table of salinity by species and zone
we might want to first check that there is an actual difference in salinity among zone
```{r}
sal_sum <- 
  catch_data_final |>
  group_by(scientific_name, zone) |>
  summarize(
    sem_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

# or with plotrix
library(plotrix)
catch_data_final |>
  group_by(scientific_name, zone) |>
  summarize(
    sem_salinity = std.error(count),
    # sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )
```

### Make point and SD Plot
```{r}
sal_sum |>
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(title = "Median Salinity by Zone with Standard Deviation Error Bars", 
       x = "Zone", 
       y = "Median Salinity (ppt)"
)
```

### Calculate mean and SE of catch data
```{r}
fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarize(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )
```


### Make point and SE plot of catch data by zone and species
```{r}
fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = zone)) +
  geom_point(position = "dodge", size = 2, pch = 23) +
  #geom_line(group = 1) +
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) +
  # add in points from master data
  # geom_point(data = catch_data_final, aes(x = zone, y = count), 
  #           position = position_jitter(width = 0.2), alpha = 0.5, size = 1, inherit.aes = FALSE) +
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_bw() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"),
   # legend.position = "bottom"
    legend.position = "none"
  ) +
  labs(
    #title = "Predatory Fish Abundance across Estuarine Salinity Gradients",
    #subtitle = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    y = "Abundance (per day)",
    fill = "Spatial Zone"
  )
```

