---
title: "WKSP_02_KeystoneExercise"
author: "Justin Proctor"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: html
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

## Module 2: R for Marine Science

## Workshop 2: **Advanced Data Wrangling**

# **Keystone exercise: Estuary Fish Survey - Data Rescue Mission**

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

### Introduction

Here we are going to wrangle in messy data on the Ross River estuary fish surveys and produce final clean tables and figure. We have 4 separate files that were provided to us by a former PhD student. We know we have data for 4 sites, 4 species, and 30 days of sampling. We also have metadata on the sites and sensor data on the environmental conditions at each site.

### Housekeeping

```{r housekeeping}
rm(list =ls())
```

### Packages

```{r packages}
library(tidyverse)
library(readxl)
```

### Load Files/Main Dataset

```{r load metadata/files}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
View(metadata)
```

```{r loading/creating main dataset}
catch1 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
```

```{r inspect individual tabs}
catch1
catch2
catch3
catch4
```

```{r bind rows together}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
rm(catch1, catch2, catch3, catch4) #clean up the environment
```

```{r alternative method for loading files}
file_path <- "../data/workshop2/estuary_catch_log.xlsx"
catch_data2 <- excel_sheets(file_path) |> 
  map_dfr(~read_excel(path = file_path, sheet = .x))
```

### Examine Data

```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

Can't tell from these outputs how many levels of site and species we have. Two options: let's check those explicitly by using distinct() or can convert to factors Let's look at the distinct values first, as we might need to clean before converting to factor vars.

```{r}
catch_data |>
  distinct(site)

catch_data |>
  distinct(species)
```

OK, so we have four species in there as well as four sites which look to be located at different points on the Ross River. Need to: - Fix site names and place in reasonable order. - Can decide this now or we can load the metadata and ensure consistency - Fix species names - to do this we'll need to make consistent and then use the taxonomic dictionary to translate to scientific names.

### Load, Inspect, and Standardize Metadata

```{r inspect metadata}
head(metadata)
glimpse(metadata)
```

site names are all over the place in metadata - will have to standardize

```{r clean up site names}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

```{r recheck}
catch_data |> 
  str()
# catch_data |> 
#   distinct(site)
glimpse(metadata)
```

### Load Taxonomic Dictionary and Sensor Data

```{r load taxonimic dictionary}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
head(species_dict)
```

```{r load the sensor data}
sensor_row <- read_csv("../data/workshop2/estuary_sonde_data.csv")
```

```{r check sensor data}
head(sensor_row)
str(sensor_row)
glimpse(sensor_row)
summary(sensor_row)
```

### Wrangle Sonde Data

```{r celan up sensor data}
sensor_data <- sensor_row |> 
  mutate(
    # parse the messy character string (day/month/year hour:minute)
    datetime = dmy_hm(timestamp),
    # convert the -999.0 hardware error to a true NA
    turbidity = na_if (turbidity, -999.0),
    # convert negative salinity values to NA
    salinity = ifelse(salinity < 0, NA, salinity),
  )
```

```{r clean up daily sensor data}
sensor_daily <- 
  sensor_data |> 
  mutate(date = as_date(floor_date(datetime, "day"))) |>
  group_by(site, date) |> 
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    mean_turbidity = mean(turbidity, na.rm = TRUE),
    .groups = "drop"
  )
```

### Notes:

-   Environmental variables needed cleaning:
-   Turbidity data has a failed value (-999)
-   Salinity had some negative values which are not possible and should be converted to NA
-   Site ok

### Wrangle Catch Data

```{r standardize variable in the catch log to match dictionaries}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_"),
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```

```{r check common names in catch log match those in the taxonomic dictionary}
catch_data |> 
  distinct(species)
species_dict |>
  distinct(common_name)
```

```{r taxonomic translation (common to scientific)}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name))
```

### Need to:

-   Join with metadata and sensor data
-   Implicit zero observations

4 x spp 4 x sites 30 surveys = 4\*4\*30

But we only have 427 rows of data - doesn't seem right. There are some days where there were zero counts of certain species at different sites. We'll need to wrangle these into a complete dataset with 480 rows, where the zero counts are explicitly represented as 0 rather than NA. This is important for our subsequent analyses and visualisations, as we want to be able to distinguish between true absences (0) and missing data (NA).

### Join Catch Data with Metadata and Sensor Data

```{r join catch data with metadata and sensor data}
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))
```

### Produce Final Table

```{r}
catch_data_final |>
  complete(
    nesting(site, date, lat, lon, zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    #convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )
```

# MAIN DATA WRANGLE DONE

## Pivot to Thinking About Ecological Hypotheses

## Differences in Environmental Variables across Zones:

### Salinity

```{r}
# Let's make a table
sal_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
   # sem_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

# And from this a plot
sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
```

Definitely differences in salinity, as we'd expect!! And we know there has been work on some of these species about responses to salinity gradients.

Let's check the other two env vars too, just to be sure.

### Turbidity

```{r}
turb_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_turbidity = sd(mean_turbidity, na.rm = TRUE),
    median_turbidity = median(mean_turbidity, na.rm = TRUE),
    .groups = "drop"
  )

turb_sum |>
  ggplot(aes(x = zone, y = median_turbidity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_turbidity - sd_turbidity, ymax = median_turbidity + sd_turbidity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Turbidity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Turbidity (NTU)")
```

### Temperature

```{r}
temp_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_temperature = sd(mean_temp, na.rm = TRUE),
    median_temperature = median(mean_temp, na.rm = TRUE),
    .groups = "drop"
  )

temp_sum |>
  ggplot(aes(x = zone, y = median_temperature)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_temperature - sd_temperature, ymax = median_temperature + sd_temperature), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Temperature by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Temperature (°C)")

```

OK so we'll continue with salinity as our main env var of interest, but we know that there are also differences in turbidity and temperature across the zones, which could also be influencing fish abundance patterns. We can discuss this in the interpretation of our results and consider potential confounding effects of these other environmental variables in our discussion.

### Publication Quality Salinity Plot

```{r}
#| fig-cap: "Salinity Gradient across zones. Ross River Estuary System, Townsville, NQ, Australia May 2026."

sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
  
```

## Patterns in Fish Abundance

### Calculate Summary Stats

```{r}
fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )
```

# Plot of Mean Catch by Zone and Species with Error Bars and Jittered Points

```{r}
fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = zone)) +
  geom_line(group = 1) +
  # add in points from master data
  geom_point(data = catch_data_final, aes(x = zone, y = count), 
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) +
  geom_point(position = "dodge", size = 4, pch = 23) +
  
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) +
  
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"),
   # legend.position = "bottom"
    legend.position = "none"
  ) +
  labs(
    #title = "Predatory Fish Abundance across Estuarine Salinity Gradients",
    #subtitle = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    y = "Abundance (per day)",
    fill = "Spatial Zone"
  )
```
